\textbf{COMP0085 Summative Assignment}

### 2023-01-02 submission.tex
update report with factor analysis and q3 convergence

%\begin{center}
%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
%    \node[main] (A) {$A$};
%    \node[main] (B) [below left of=A] {$B$};
%    \node[main] (C) [below right of=A] {$C$};
%    \node[main] (E) [below right of=C] {$E$};
%    \node[main] (D) [below left of=C] {$D$};
%    \node[main] (H) [below of=E] {$H$};
%    \node[main] (G) [right of=E] {$G$};
%    \node[main] (I) [below left of=H] {$I$};
%    \node[main] (F) [above left of=I] {$F$};
%
%    \draw (A) -- (B);
%    \draw (A) -- (C);
%    \draw (B) -- (F);
%    \draw (B) -- (D);
%    \draw (C) -- (D);
%    \draw (C) -- (E);
%    \draw (E) -- (G);
%    \draw (G) -- (H);
%    \draw (D) -- (F);
%    \draw (E) -- (H);
%    \draw (F) -- (H);
%    \draw (F) -- (I);
%    \draw (H) -- (I);
%
%    \draw[dashed] (I) -- (E);
%    \draw[dashed] (F) -- (E);
%    \draw[dashed] (C) -- (F);
%    \draw[dashed] (B) -- (D);
%    \draw[dashed] (B) -- (C);
%
%\end{tikzpicture}
%\end{center}

@@

where the dashed lines are edges added to triangulate the moralised graph.

@@

%\begin{center}
%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
%    \node[main] (ABC) {$ABC$};
%    \node[square] (BC) [square,draw] [below of=ABC] {$BC$};
%    \node[main] (BCDF) [below of=BC] {$BCDF$};
%    \node[square] (CF) [square,draw] [below of=BCDF] {$CF$};
%    \node[main] (CEF) [below of=CF] {$CEF$};
%    \node[square] (EF) [square,draw] [below of=CEF] {$EF$};
%    \node[main] (EFHI) [below of=EF] {$EFHI$};
%    \node[square] (EH) [square,draw] [below of=EFHI] {$EH$};
%    \node[main] (EGH) [below of=EH] {$EGH$};
%    \draw (ABC) -- (BC);
%    \draw (BC) -- (BCDF);
%    \draw (BCDF) -- (CF);
%    \draw (CF) -- (CEF);
%    \draw (CEF) -- (EF);
%    \draw (EF) -- (EFHI);
%    \draw (EFHI) -- (EH);
%    \draw (EH) -- (EGH);
%\end{tikzpicture}
%\end{center}

\begin{center}
\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
    \node[main] (ABC) {$ABC$};
    \node[main] (BCDF) [below of=ABC] {$BCDF$};
    \node[main] (CFH) [below of=BCDF] {$CFH$};
    \node[main] (FHI) [below left of=CFH] {$FHI$};
    \node[main] (CEH) [below right of=CFH] {$CEH$};
    \node[main] (EGH) [below of=CEH] {$EGH$};
    \draw (ABC) -- (BCDF);
    \draw (BCDF) -- (CFH);
    \draw (CFH) -- (FHI);
    \draw (CFH)  -- (CEH);
    \draw (EGH)  -- (CEH);
\end{tikzpicture}
\end{center}

where the circular nodes are cliques.

@@

@@

@@

Plotting the convergence of the variational approximation for different $\sigma$'s:

@@

We can see that when $\sigma$ is smaller, we are able to converge at a higher free energy. However, convergence is only reached after more steps, as seen in the plot of $\log(F(t)-F(t-1))$

>>>>

@@

where the dashed lines are edges added to triangulate the moralised graph.
\newpage

@@

%\begin{center}
%\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
%    \node[main] (ABC) {$ABC$};
%    \node[main] (BCDF) [below of=ABC] {$BCDF$};
%    \node[main] (CFH) [below of=BCDF] {$CFH$};
%    \node[main] (FHI) [below left of=CFH] {$FHI$};
%    \node[main] (CEH) [below right of=CFH] {$CEH$};
%    \node[main] (EGH) [below of=CEH] {$EGH$};
%    \draw (ABC) -- (BCDF);
%    \draw (BCDF) -- (CFH);
%    \draw (CFH) -- (FHI);
%    \draw (CFH)  -- (CEH);
%    \draw (EGH)  -- (CEH);
%\end{tikzpicture}
%\end{center}

@@

\begin{center}
\begin{tikzpicture}[node distance={10mm}, main/.style = {draw, circle}, square/.style={fill=black, regular polygon,regular polygon sides=4}]
    \node[main] (A) {$A$};
    \node[square] (ABC) [square,draw] [below of=A];
    \node[main] (B) [below left of=ABC] {$B$};
    \node[main] (C) [below right of=ABC] {$C$};
    \node[square] (BCDF) [square,draw] [below left of=C];
    \node[square] (CEH) [right of=C];
    \node[main] (E) [above right of=CEH] {$E$};
    \node[square] (GEH) [below right of=E];
    \node[main] (G) [right of=GEH] {$G$};
    \node[main] (D) [below left of=BCDF] {$D$};
    \node[square] (CFH) [below right of=C];
    \node[main] (H) [right of=CFH] {$H$};
    \node[square] (FHI) [below left of = H];
    \node[main] (I) [below right of=FHI] {$I$};
    \node[main] (F) [below right of=BCDF] {$F$};

    \draw (ABC) -- (A);
    \draw (ABC) -- (C);
    \draw (ABC) -- (B);
    \draw (BCDF) -- (B);
    \draw (BCDF) -- (C);
    \draw (BCDF) -- (D);
    \draw (BCDF) -- (F);
    \draw (CEH) -- (C);
    \draw (CEH) -- (E);
    \draw (CEH) -- (H);
    \draw (GEH) -- (G);
    \draw (GEH) -- (E);
    \draw (GEH) -- (H);
    \draw (CFH) -- (C);
    \draw (CFH) -- (F);
    \draw (CFH) -- (H);
    \draw (FHI) -- (F);
    \draw (FHI) -- (H);
    \draw (FHI) -- (I);
\end{tikzpicture}
\end{center}
\newpage

@@

Using our factor analysis model, we can describe the biochemical pathway as:

\[\delta[\textbf{x}] = \Lambda \textbf{z} + \epsilon\]

where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$, and the latent factors $z \sim \mathcal{N} (0, I)$.

From the graph structure, we know that:

\[\Lambda = \begin{bmatrix}
                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
                 \Lambda_{BA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
                 \Lambda_{CA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & \Lambda_{DC} & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & \Lambda_{EC} & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & \Lambda_{FB} & 0 & \Lambda_{FD} & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & 0 & 0 & \Lambda_{HE} & 0 & \Lambda_{HG} & 0 & 0  \\
                 0 & 0 & 0 & 0 & 0 & \Lambda_{IF} & 0 & \Lambda_{IH} & 0 \\
         \end{bmatrix} \text{ and }
    \textbf{z} = \begin{bmatrix}
                 z_A  \\
                 z_B  \\
                 z_C  \\
                 z_D  \\
                 z_E  \\
                 z_F  \\
                 z_G  \\
                 z_H  \\
                 z_I \\
         \end{bmatrix}
\]

Having observations for $\delta[B]$, $\delta[D]$, $\delta[E]$ and $\delta[G]$:

\[
\begin{bmatrix}
                 \delta[B]  \\
                 \delta[D]  \\
                 \delta[E]  \\
                 \delta[G]  \\
         \end{bmatrix} = \begin{bmatrix}
                 \Lambda_{BA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & \Lambda_{DC} & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & \Lambda_{EC} & 0 & 0 & 0 & 0 & 0 & 0  \\
                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
         \end{bmatrix} \begin{bmatrix}
                 z_A  \\
                 z_B  \\
                 z_C  \\
                 z_D  \\
                 z_E  \\
                 z_F  \\
                 z_G  \\
                 z_H  \\
                 z_I \\
         \end{bmatrix} +  \begin{bmatrix}
                 \epsilon_B  \\
                 \epsilon_D  \\
                 \epsilon_E  \\
                 \epsilon_G  \\
         \end{bmatrix}
\]

We can see that these simplify to the equations:

\begin{gather*}
    \delta[B] = \Lambda_{BA}z_A + \epsilon_B\\
    \delta[D] = \Lambda_{DC}z_C  + \epsilon_D\\
    \delta[E] = \Lambda_{EC}z_C  + \epsilon_E\\
    \delta[G] = \epsilon_G\\
\end{gather*}

Thus, we see that the only latent variables present are $z_A$ and $z_C$, so would expect to recover the factors of $A$ and $C$, the two parent nodes of the observations.

@@

Plotting the free energy at each partial expectation step of the variational approximation for different $\sigma$'s:

@@

We know that our free energy is a upper bounded on the log likelihood:

\[\log P(\mathcal{X} | \theta) \geq \mathcal{F}(q, \theta)\]

In the variational expectation step, $\log P(\mathcal{X} | \theta)$ is fixed and we adjust our approximation $q$ to reach this upper bound. We know that $\sigma$ quantifies the noise of $\textbf{x}$, thus a higher $\sigma$ means a wider spread in our distribution $\log P(\mathcal{X} | \theta)$, meaning we are reducing our upper bound for $\mathcal{F}(q, \theta)$. As such, we can see in the plot for free energy above that when $\sigma$ is increased, our free energy converges to a lower value, due to being bounded above by a lower log-likelihood. Moreover, by reducing the upper bound, we see in the plot of $\log(F(t)-F(t-1))$ that our free energy is able to converge faster. Because we have reduced the upper bound by increasing $\sigma$, our free energy can reach this upper bound faster.
